Detecting the impact of land cover change on observed rainfall.

### 2019-03-14 Manuscripts/PeerJRainfall_Landuse/Rainfall_Landuse.tex
paper for submission

\usepackage{lipsum} \usepackage{textcomp}

@@

\begin{abstract}
Analysis of observational data to identify relationships between
rainfall and land cover change are difficult due to multiple
environmental factors that cannot be strictly controlled. In this study
we present a methodology to investigate the relationship using
statistical methods on data from best available sources at two sites in
Australia. Gridded data of rainfall and tree cover were used as
spatially corresponding local conditions. Large scale effects were
represented by climatic indicators, such as SOI and IOD. Regression
analysis and step trend tests were used to assess the effect of abrupt
land surface intervention. At a Queensland site, significant tree cover
change between 2002 - 2005 did not result in strong statistically
significant precipitation changes. On the other hand, results from a
bushfire affected NSW/VIC region suggests significant changes in the
rainfall. This indicates the method works better when a abrupt change in
the data can be clearly identified. The results from the step trend test
implied a positive relationship between the tree cover and the rainfall
at 0.1 significance level in both locations in data up to 2009. However,
high rainfall variability and possible regrowth meant that no signifcant
changes were observed in alonger time sereies to 2015.
% Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text.
\end{abstract}

@@

Land use and land cover changes can lead to changes in the local
climate. Empirical and modelling studies have found that cloud types and
rainfall are correlated to large scale vegetation cover changes, such as
deforestation in the Amazon and in the Sahel (Chagnon and Bras 2005;
Pinto et al. 2009; Wang et al. 2009; Mei and Wang 2010; Kucharski, Zeng,
and Kalnay 2013; Pitman and Lorenz 2016) and afforestation in south
Israel (Otterman et al. 1990; Ben-Gai et al. 1998). Using airborne
measurement in Western Australia, Junkermann et al. (2009) showed a
significantly higher level of aerosols over an agricultural area
compared to an adjacent natural vegetation. They suggested that a
modification of aerosol concentrations due to deforestation could have
contributed to a reduction of local rainfall, as more, but smaller rain
droplets were observed. Nair et al. (2011) reported from the Bunny Fence
Experiment in Western Australia that local land use change altered the
synoptic west coast trough dynamics and surface roughness, and this
resulted in an observed rainfall decrease. Maximum temperatures were
also found to be sensitive to land cover change in eastern Australia
(McAlpine et al. 2007).

@@

In Australia, significant tree cover change has mainly occurred in the
north east and south east of the continent, as well as in the southwest
of Western Australia. According to the National Dynamic Land Cover
Dataset (DLCD) (Lymburner et al. 2010), most of these areas have
experienced decreases in the Enhanced Vegetation Index (EVI) post 2000
aws derived from satellite data. As an index for vegetation greenness,
the decreasing values indicate lower biomass over time in the tree cover
regions. The possible EVI reduction might be due to land clearing, bush
fires or drought.

@@

Two regions were selected where significant tree cover change since 2000
was reported. The first region is located in south central Queensland
partly covering the north of the Murray Darling Basin (MDB) (site 1 in
Figure @ref(fig:selreg)). High rates of land clearing have been reported
in this region during the early 2000s (Department of Natural Resources
and Water 2007). The second study region is located at the border of New
South Wales and Victoria, and includes the Snowy Mountain ranges (site 2
in Figure @ref(fig:selreg)). Severe bush fires occurred in this area in
early 2003 (see Figure @ref(fig:figure2bushfire)). The 2003 bush fires
were the largest and the worst in the last 60 years (The State
Government of Victoria 2011). Two thirds of Kosciuszko national park was
heavily burned and regrowth was reported to be slow due to drought and
cold conditions (ABC News 2003), and the type of species in this region.
However, in the longer term, after an early high transpiration period a
recovery of pre-fire evapotranspiration would be expected (Kuczera
1987). For the purpose of this study, significant tree cover loss has
happened in both study areas in the last decade, either permanently or
temporarily.

\begin{figure}
\includegraphics[width=0.9\linewidth]{figures/bushfire_nswvic} \caption{Location of bushfires occurring in January 2003, in and around the NSW/VIC study region, as shown by the red pixels. The map shows the large area in the Kosciuszko national park that has been burned. Some locations in the southwest of teh Australian Capital Territory (ACT) have also experienced intensive bushfires.}\label{fig:figure2bushfire}
\end{figure}

@@

Due to the difference in nature of the land cover change in the two
regions, the post-change vegetation status is hypothesized to be
different as well (see Figure @ref(fig:figure3-tc-simple)).

@@

\hypertarget{Data}{\subsection{Data}\label{Data}}

Several land surface data sets were used in this study. The main one was
the MOD44B product Global Vegetation Continuous Field data set (version
5). This data set provides estimates of percent tree cover (percentage
of ground surface covered by trees) at a grid resolution of 250 m
(Townshend et al. 2011), which is finer then the earlier mentioned
burned product MCD45A1. The data set is available on an annual time
interval for the study period of 2000 - 2015. The tree cover data was
produced from 16-day Terra MODIS Land Surface Reflectance data and Land
Surface Temperature (Townshend et al. 2011). The National Dynamic Land
Cover Dataset (DLCD) (Lymburner et al. 2010) from the Australian
Collaborative Land Use Mapping Program (ACLUMP) was used to verify the
trend of vegetation cover change calculated from the previous data set.
This data set, developed by Geoscience Australia and Australian Bureau
of Agricultural and Resource Economics and Sciences (ABARES), is the
first nationally consistent and thematically comprehensive land cover
reference for Australia. The DLCD is based on the 16-day Enhanced
Vegetation Index (EVI), again from the MODIS satellite, between April
2000 and April 2015. It also has a grid resolution of 250 m. The data
set provides information on the final land cover types (as in 2015) and
estimated trend of EVI statistics (annual mean, maximum and minimum).

@@

\subsection{Statistical method}\label{statistical-method}

@@

To assess the actual causal relationship between the tree cover and the
rainfall a flexible regression model is applied (discussed in detail
below). A step change is not directly obvious in the time series of the
rainfall anomalies (Figure @ref(fig:ts-mean)) for both regions, even
though the data is deseasonalised and detrended. In this study we apply
different statistical methods to analyse the effect of tree cover change
on rainfall. Both methods make use of a regression model to remove
year-on-year variability in rainfall to strengthen the tree cover change
signal.

@@

\subsubsection{Regression model}\label{reg_model}

As highlighted in the introduction, the Australian climate is influenced
by sea surface temperatures in the tropical Pacific and Indian Oceans,
as well as pressure systems in the Southern Ocean (BoM 2012b). Risbey et
al. (2009) compared five large-scale drivers, including ENSO (measured
by SOI and the Tropical Pacific Sea surface temperatures (SSTs)), IOD,
SAM, MJO (Madden-Julian oscillation) and blocking, in relation to
Australian rainfall variability. The MJO is a large scale
eastward-propagating wave-like disturbance located around equatorial
latitudes (Risbey et al. 2009). They identified SOI as the most
important index among all climate indices tested for broad parts of
Australia (including QLD and NSW/VIC) in almost any season. In this
study, four climate indices were selected from the main climatic
indicators (see section \protect\hyperlink{Data}{Data}) and used as the
explanatory variables in the model for each study region. A further
complicating factor is the influence of the ``millenium drought'' over
the study period and in particular the change to wet conditions in 2010
- 2011 (Dijk and Viney 2013). Therefore, the spatially averaged monthly
rainfall in the Murray Darling Basin (MDB, downloaded from
\href{http://www.bom.gov.au/web01/ncc/www/cli_chg/timeseries/rain/allmonths/mdb/latest.txt}{the
Bureau of Meteorology}) was used to explain the year-on-year variation
in the rainfall in the regions. As climate alone are unlikely able to
fully explain the temporal variation in rainfall in the regions,
including the drought (Dijk and Viney 2013; Westra and Sharma 2010),
this variable was included. Since both regions at least partly overlap
with the MDB, the average rainfall across was assumed to be a useful
explaining variable.

Correlations between rainfall and each climate index were analysed.
Rainfall in each study region was first deseasonalised and detrended
using the seasonal decomposition function \texttt{ds} in the package
\texttt{deseasonalise} in R (R Core Team, 2018). Using detrended data
gives a better indication of the underlying correlation compared to the
correlation between trends in the data (Smith and Timbal 2012). The
cross-correlations between the deseasonalised and detrended rainfall and
the climatic indices were tested using the Pearson's product moment
correlation method, assuming the relationships are linear. Although the
optimal technique for exploring the correlation with each index could be
different as described in Risbey et al. (2009), the Pearson's method was
applied to all indices for consistency. Because the PDO describes the
multi-decadal SST with lower frequency (MacDonald and Case 2005;
Zanchettin et al. 2008; Kamruzzaman, Beecham, and Metcalfe 2011),
instead of 37-year rainfall data, a longer period (115 years, from 1900
to 2015) was used to estimate the correlation with PDO, up to lag 24.
For the other indices, the 37-year data was used.

@@

Based on the correlation between the climatic indices and rainfall in
the regions (as shown in Figure @ref(fig:cor-rain-qld) and Figure
@ref(fig:cor-rain-nsw)), it can be concluded that:

\begin{itemize}
  \setlength{\itemsep}{0cm}
  \setlength{\parskip}{0cm}
  \item In QLD, the correlation between rainfall and SOI at zero time lags is the strongest across all indices, outweighing the other ENSO indicators. IOD and PDO have a weak influence in QLD.
  \item In NSW/VIC, again the SOI has the strongest correlation with rainfall, followed by the IOD. For both, the strongest correlations occur at the zero time lags. 
  \item in both cases PDO had the weakest correlations, and this factor was not further considered as a predictor.
  \item in general, for the better correlated indices, strongest correlations occured at zero time lags.
\end{itemize}

@@

We assumed all the factors are additive smooth components in determining
rainfall following Kamruzzaman, Beecham, and Metcalfe (2011). In this
case, the rainfall model is a generalised additive model (GAM) (Hastie
and Tibshirani 1986) with a log link function \texttt{g()} and assuming
the residuals are gamma distributed (see Figure @ref(fig:hist-rain)).
This means all predictors are modelled as smooth functions, in this case
using the shrinkage version of the cubic regression splines (Wood 2011).
All splines were limited to 3 knots in flexibility (Wood 2011), to
reduce the risk of overfitting. \vspace{0.5cm}

\begin{equation}
\begin{array}{lll}
g(E(\mathbf{R}_r)) = &\beta_0 + s_1(\mathbf{MDB_{monthlyRain}} + \\ &s_2(\mathbf{SOI}) + s_3(\mathbf{IOD}) + \\                &s_4(\mathbf{Nino3.4}) + s_5(\mathbf{Nino4}) + s_6(\mathbf{Season}) + \\
           &\beta_1\mathbf{Trend} + \boldsymbol{\epsilon}_r
\end{array}
(\#eq:model)
\end{equation}

The bold letters represent the time series vectors. The region is
indicated by \(r\), while \(\beta_u\) (\(_u\)=0, 1) are the fitted
coefficients in the model. \(s_v\) (\(_v\)=1, 2, 3,\ldots{}) are the
smooth penalized cubic regression spline functions on the climatic
indices and the season. MDB\_\{monthlyRain\} is the spatially averaged
monthly rainfall across the Murray Darling Basin. Apart from dropping
PDO as a predictor, all other climate indices were included, allowing
the model to select the appropriate predictors.

@@

One of the main difficulties in empirical observation studies related to
the effect of land cover change on rainfall is the lack of continuous
monitoring of land surface variables. Moreover, no specific variable can
possibly be defined that can clearly represent the land surface process.
Given the lack of a full picture of the land surface process, a factor
variable was used in the regression model to represent the abrupt land
surface change (see Equation @ref(eq:model)). The change could be a
result of either land clearing or bush fires as long as it is permanent
or takes a long time to recover. As indicated we approached the problem
in two different models ways.

In the first method, the tree cover change was used as a predictor in
the regression model, represented by a factor variable \textbf{LC}. The
significance of the coefficient of \textbf{LC}, denoted as \(\beta'_5\)
in Equation @ref(eq:model), can be determined by a ratio test.

\begin{equation}
  \mbox{LC} = \left\{
  \begin{array}{ll}
     \mbox{Trees} \\
     \mbox{Removed}
  \end{array} \right.\\
  (\#eq:dummyC)
\end{equation}

@@

\begin{equation}
\begin{array}{lll}
g(E(\mathbf{R}_r)) = &\beta'_0 + s'_1(\mathbf{MDB_{monthlyRain}} + \\
  &s'_2(\mathbf{SOI}) + s'_3(\mathbf{IOD}) + \\
  &s'_4(\mathbf{Nino3.4}) + s'_5(\mathbf{Nino4}) + s'_6\mathbf{Season} + \\
  &\beta'_1\mathbf{Trend} + \beta'_2\mathbf{LC} + \boldsymbol{\epsilon'}_r
\end{array}
  (\#eq:model2)
\end{equation}

One of the difficulties is to point an exact time to the changes in the
vegetation cover in the two regions. In the QLD region, no exact time
can be assigned to the land clearing. According to the SLATS reports,
the most substantial clearing occurred between 2003 - 2004 . However,
the information on the change in type of land cover during the time
period is missing. Therefore, four scenarios were initially tested in
the analysis (\textbf{see supplementary material}). In the NSW/VIC
region, severe bush fires were reported in early January 2003. Hence the
``tree'' cover state was up to December 2002 then it was changed to
``removed'' state from January 2003. As a starting date, the regression
model was run from 1979 for both regions.

@@

To support the regression analysis, a step trend test was used to detect
changes in rainfall as a result of vegetation cover change. This
nonparametric statistical test was modified from the Mann-Whitney
Rank-Sum test by (Hirsch and Gilroy 1985) and can identify a step change
in data which is cross correlated. The gridded rainfall dataset used in
this study has a high spatial correlation between neighbouring pixels
due to the underlying interpolation method. The advantages of using the
Rank-Sum test therefore are: (1) it does not depend on assumptions of
the data distribution; (2) it is not restricted to datasets with no
missing data; (3) it is robust and not as easily influenced by outliers
and negative numbers (Hirsch and Gilroy 1985).

The rainfall residuals from the regression model in Equation
@ref(eq:model2) were used for the step trend test. According to (Hirsch
and Gilroy 1985), to detect a step change, using deseasonalised and
detrended data is important. Furthermore, as rainfall can only be
partially attributed to local sources and conditions, other effects are
introduced by large scale dynamics and changes in other climatic
factors. The assumption is that the regression model should remove these
effects, and additionally deseasonalise and detrend the rainfall data.
As a result, the local landuse effects are amplified in the variation of
the remaining residuals. The test, described in the following sub
section, subsequently associates trends in the rainfall residuals with
tree cover changes.

@@

As indicated, the step trend test is a modified version of the
Mann-Whitney rank-sum statistic (Hirsch and Gilroy 1985). As a
nonparametric rank-based test, the Mann-Whitney test does not use the
exact values of rainfall but depends on the ranks of the data. For each
month, rainfall residuals of each year were ranked in an ascending
order. The ranking of January rainfall in a sample pixel k in QLD is
illustrated below:

\begin{longtable}[]{@{}ccc@{}}
\caption{Example of ranking rainfall residuals}\tabularnewline
\toprule
Year & Rainfall residuals & Rank \(R'_{1k}\)\tabularnewline
\midrule
\endfirsthead
\toprule
Year & Rainfall residuals & Rank \(R'_{1k}\)\tabularnewline
\midrule
\endhead
1998 & -0.3 & 6\tabularnewline
1999 & -60.9 & 2\tabularnewline
2000 & -16.1 & 4\tabularnewline
2001 & -71.7 & 1\tabularnewline
2002 & 111.1 & 7\tabularnewline
2005 & -7.2 & 5\tabularnewline
2006 & -60.5 & 3\tabularnewline
\bottomrule
\end{longtable}

@@

The before and after period in the data formed two groups of samples.
The split point of the two periods was based on the timing of the
vegetation cover changes. In the QLD region, changes occurred anytime
during 2003 and 2004. In contrast to the previous method, the time
period covering the land cover change was excluded, as the nonparametric
test allows missing data. Hirsch and Gilroy (1985) also pointed out that
the power of the test is higher if the data of the change period is
ignored. Hence 2003 and 2004 were excluded from the analysis. As a
result, the after-change period was 2005 - 2015 for the Queensland
location.

In the case of NSW/VIC, the bushfires broke out in early January 2003.
The change was within a relatively short period of the year. Therefore
the after change period in this region still started in January 2003.
Following Hirsch and Gilroy (1985), the before period was set to five
years (1998 - 2002) in both regions. The length of the after period is
difficult for the NSW/VIC as the regrowth would at some point have
impacted the local effects.

There is a further complication in the data. The year 2009 was the final
year of the millenium drought (Dijk and Viney 2013), followed by two
very wet years (also visible in Figure @ref(fig:ts-mean) in 2010 and
2011. As highlighted earlier, this means the the length of after period
could influence the analysis by the inclusion of more or fewer years. as
a result the period up to (and including) 2009 was compared to the full
period up to 2015. Further analysis up to 2010 is included in the
supplementary material.

@@

\begin{equation}
  W_{jk} = \sum_{i=1}^{n_1}R'_{ijk}.
  (\#eq:Wj)
\end{equation}

@@

Instead of completing the Mann-Whitney U-test, Hirsch and Gilroy (1985)
applied a standard normal Z test to the rank-sum statistics. As
highlighted, this modified test accounts for serial and cross
correlation in the data. In the case here, the deseasonalised and
detrended data shows little autocorrelation in the time series but
possesses strong cross correlation between neighbouring pixels, i.e.
\(R>0.99\).

@@

\begin{equation}
  Z'=\frac{\sum_{j=1}^{12}\sum_{k=1}^{ns}W_{jk}-12\cdot ns\cdot\mu_w}{\sqrt{Var(\sum_{j=1}^{12}\sum_{k=1}^{ns}W_{jk})}}.
  (\#eq:Z)
\end{equation}

@@

The null hypothesis (\(H_0\)) in this study is that there was no change
in rainfall due to land surface intervention. The results of the step
trend test can be interpreted according to the sign of the Z' score (see
Table @ref(tab:Zscore) (Chapter 23, P887 Hipel and McLeod 1994)). Z' is
normally distributed similar to the standard normal statistics Z. Hence
it can be compared to a standard normal distribution to determine the p
value.

>>>>

\usepackage{lipsum} \usepackage{textcomp, rotating}

@@

\begin{abstract}
Analysis of observational data to pinpoint impact of land cover change
on local rainfall are difficult due to multiple environmental factors
that cannot be strictly controlled. In this study we use a statistical
approach to identify the relationship between removal of tree cover and
rainfall with data from best available sources for two large areas in
Australia. Gridded rainfall data between 1979 and 2015 was used for the
areas, while large scale (exogenous) effects were represented by mean
rainfall across a much larger area and climatic indicators, such as SOI
and IOD. Both general additive modelling and step trend tests were used
for the analysis. For a region in south central Queensland, the reported
change in tree clearing between 2002 - 2005 did not result in strong
statistically significant precipitation changes. On the other hand,
results from a bushfire affected region on the border of New SOuth Wales
and Victoria suggests significant changes in the rainfall due to changes
in tree cover. This indicates the method works better when a abrupt
change in the data can be clearly identified. The results from the step
trend test also mainly identified a positive relationship between the
tree cover and the rainfall at p \textless{} 0.1 at the NSW/Victoria
region. High rainfall variability and possible regrowth could have
impacted the results in the Queensland region.
% Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text.
\end{abstract}

@@

Land use and land cover changes can lead to changes in the local
climate. Empirical and modelling studies have found that cloud types and
rainfall are correlated to large scale vegetation cover changes, such as
deforestation in the Amazon and in the Sahel (Chagnon and Bras 2005;
Pinto et al. 2009; Wang et al. 2009; Mei and Wang 2010; Kucharski, Zeng,
and Kalnay 2013; Pitman and Lorenz 2016) and afforestation in south
Israel (Otterman et al. 1990; Ben-Gai et al. 1998). Using airborne
measurements in Western Australia, Junkermann et al. (2009) showed a
significantly higher level of aerosols over an agricultural area
compared to an adjacent area with natural vegetation. They suggested
that a modification of aerosol concentrations due to deforestation could
have contributed to a reduction of local rainfall, as more, but smaller
rain droplets were observed. Nair et al. (2011) reported from the Bunny
Fence Experiment in Western Australia that local land use change altered
the synoptic west coast trough dynamics and surface roughness, and this
resulted in an observed rainfall decrease. Maximum temperatures were
also found to be sensitive to land cover change in eastern Australia
(McAlpine et al. 2007).

@@

In Australia, significant tree cover change has mainly occurred in the
north east and south east of the continent, as well as in the southwest
of Western Australia. According to the National Dynamic Land Cover
Dataset (DLCD) (Lymburner et al. 2010), most of these areas have
experienced decreases in the Enhanced Vegetation Index (EVI) post 2000,
as derived from satellite data. As an index for vegetation greenness,
the decreasing values of EVI indicate lower biomass over time in the
tree cover regions. The possible EVI reduction might be due to land
clearing, bush fires or drought.

@@

Two regions were selected where significant tree cover change since 2000
was reported. The first region is located in south central Queensland
(QLD) partly covering the north of the Murray Darling Basin (MDB) (site
1 in Figure \ref{fig:selreg}). High rates of land clearing have been
reported in this region during the early 2000s (Department of Natural
Resources and Water 2007). The second study region is located at the
border of New South Wales and Victoria (NSW/VIC), and includes the Snowy
Mountain ranges (site 2 in Figure \ref{fig:selreg}). Severe bush fires
occurred in this area in early 2003 (see Figure \ref{fig:bushfire}). The
2003 bush fires were the largest and the worst in the last 60 years (The
State Government of Victoria 2011). Two thirds of Kosciuszko national
park was heavily burned and regrowth was reported to be slow due to
drought and cold conditions (ABC News 2003), and the type of species in
this region. However, in the longer term, after an early high
transpiration period a recovery of pre-fire evapotranspiration would be
expected (Kuczera 1987). For the purpose of this study, significant tree
cover loss has happened in both study areas in the last decade, either
permanently or temporarily.

\begin{figure}
\includegraphics[width=0.9\linewidth]{figures/bushfire_nswvic} \caption{Location of bushfires occurring in January 2003, in and around the NSW/VIC study region, as shown by the red pixels. The map shows the large area in the Kosciuszko national park that has been burned. Some locations in the southwest of the Australian Capital Territory (ACT) have also experienced intensive bushfires.}\label{fig:bushfire}
\end{figure}

@@

Due to the difference in nature of the land cover change in the two
regions, the post-change vegetation status is hypothesized to be
different as well (see Figure \ref{fig:figure3-tc-simple}).

@@

\hypertarget{Data}{\section{Data}\label{Data}}

Several land surface data sets were used in this study. The main one was
the MOD44B product Global Vegetation Continuous Field data set (version
5). This data set provides estimates of percent tree cover (percentage
of ground surface covered by trees) at a grid resolution of 250 m
(Townshend et al. 2011), which is finer than the earlier mentioned
burned product MCD45A1. The data set is available on an annual time
interval for the study period of 2000 - 2015. The tree cover data was
produced from 16-day Terra MODIS Land Surface Reflectance data and Land
Surface Temperature (Townshend et al. 2011). The National Dynamic Land
Cover Dataset (DLCD) (Lymburner et al. 2010) from the Australian
Collaborative Land Use Mapping Program (ACLUMP) was used to verify the
trend of vegetation cover change calculated from the previous data set.
This data set, developed by Geoscience Australia and Australian Bureau
of Agricultural and Resource Economics and Sciences (ABARES), is the
first nationally consistent and thematically comprehensive land cover
reference for Australia. The DLCD is based on the 16-day Enhanced
Vegetation Index (EVI), again from the MODIS satellite, between April
2000 and April 2015. It also has a grid resolution of 250 m. The data
set provides information on the final land cover types (as in 2015) and
estimated trend of EVI statistics (annual mean, maximum and minimum).

@@

\section{Statistical method}\label{statistical-method}

@@

To assess the actual causal relationship between the tree cover and the
rainfall a flexible regression model is applied (discussed in detail
below). A step change is not directly obvious in the time series of the
rainfall anomalies (Figure \ref{fig:ts-mean}) for both regions, even
though the data is deseasonalised and detrended. In this study we apply
different statistical methods to analyse the effect of tree cover change
on rainfall. Both methods make use of a regression model to remove
year-on-year variability in rainfall to strengthen the tree cover change
signal.

@@

\subsection{Regression model}\label{reg_model}

As highlighted in the introduction, the Australian climate is influenced
by sea surface temperatures in the tropical Pacific and Indian Oceans,
as well as pressure systems in the Southern Ocean (BoM 2012b). Risbey et
al. (2009) compared five large-scale drivers, including ENSO (measured
by SOI and the Tropical Pacific Sea surface temperatures (SSTs)), IOD,
SAM, MJO (Madden-Julian oscillation) and blocking, in relation to
Australian rainfall variability. The MJO is a large scale
eastward-propagating wave-like disturbance located around equatorial
latitudes (Risbey et al. 2009). They identified SOI as the most
important index among all climate indices tested for broad parts of
Australia (including QLD and NSW/VIC) in almost any season. In this
study, four climate indices were selected from the main climatic
indicators (see section \protect\hyperlink{Data}{Data}) and used as the
explanatory variables in the model for each study region. A further
complicating factor is the influence of the ``millenium drought'' over
the study period and in particular the change to wet conditions in 2010
- 2011 (Dijk and Viney 2013). Therefore, the spatially averaged monthly
rainfall in the Murray Darling Basin (MDB, downloaded from
\href{http://www.bom.gov.au/web01/ncc/www/cli_chg/timeseries/rain/allmonths/mdb/latest.txt}{the
Bureau of Meteorology}) was used to explain the year-on-year variation
in the rainfall in the regions. As climate alone are unlikely able to
fully explain the temporal variation in rainfall in the regions,
including the drought (Dijk and Viney 2013; Westra and Sharma 2010),
this variable was included. Since both regions at least partly overlap
with the MDB, the average rainfall for the entire basin was assumed to
be a useful explaining variable.

Correlations between rainfall and each climate index were analysed.
Rainfall in each study region was first deseasonalised and detrended
using the seasonal decomposition function \texttt{ds} in the package
\texttt{deseasonalise} in R (R Core Team, 2018). Using detrended data
gives a better indication of the underlying correlation by removing the
underlying correlation in the data (Smith and Timbal 2012). The
cross-correlations between the deseasonalised and detrended rainfall and
the climatic indices were tested using the Pearson's product moment
correlation method, assuming the relationships are linear. Although the
optimal technique for exploring the correlation with each index could be
different as described in Risbey et al. (2009), the Pearson's method was
applied to all indices for consistency. Because the PDO describes the
multi-decadal SST with lower frequency (MacDonald and Case 2005;
Zanchettin et al. 2008; Kamruzzaman, Beecham, and Metcalfe 2011),
instead of 37-year rainfall data, a longer period (115 years, from 1900
to 2015) was used to estimate the correlation with PDO, up to lag 24.
For the other indices, the 37-year data was used.

@@

Based on the correlation between the climatic indices and rainfall in
the regions (as shown in Figure \ref{fig:cor-rain-qld} and Figure
\ref{fig:cor-rain-nsw}), it can be concluded that:

\begin{itemize}
  \setlength{\itemsep}{0cm}
  \setlength{\parskip}{0cm}
  \item In QLD, the correlation between rainfall and SOI at zero time lags is the strongest across all indices, outweighing the other ENSO indicators. IOD and PDO have a weak influence in QLD.
  \item In NSW/VIC, again the SOI has the strongest correlation with rainfall, followed by the IOD. For both, the strongest correlations occur at the zero time lags. 
  \item In both cases PDO had the weakest correlations, and this factor was not further considered as a predictor.
  \item In general, for the better correlated indices, strongest correlations occured at zero time lags.
\end{itemize}

@@

We assumed all the factors are additive smooth components in determining
rainfall following Kamruzzaman, Beecham, and Metcalfe (2011). In this
case, the rainfall model is a generalised additive model (GAM) (Hastie
and Tibshirani 1986) with a log link function \texttt{g()} and assuming
the residuals are gamma distributed (see Figure \ref{fig:hist-rain}).
This means all predictors are modelled as smooth functions, in this case
using the shrinkage version of the cubic regression splines (Wood 2011).
All splines were limited to 3 knots in flexibility (Wood 2011), to
reduce the risk of overfitting. \vspace{0.5cm}

\begin{equation}
\begin{array}{lll}
g(E(\mathbf{R}_r)) = &\beta_0 + s_1(\mathbf{MDB_{monthlyRain}} + s_2(\mathbf{SOI}) + s_3(\mathbf{IOD}) + \\
&s_4(\mathbf{Nino3.4}) + s_5(\mathbf{Nino4}) + s_6(\mathbf{Season}) + \\
           &\beta_1\mathbf{Trend} + \boldsymbol{\epsilon}_r
\end{array}
\label{eq:model}
\end{equation}

The bold letters represent the time series vectors. The region is
indicated by \(r\), while \(\beta_u\) (\(_u\)=0, 1) are the fitted
coefficients in the model. \(s_v\) (\(_v\)=1, 2, 3,\ldots{}) are the
smooth penalized cubic regression spline functions on the climatic
indices and the season. MDB\textsubscript{monthlyRain} is the spatially
averaged monthly rainfall across the Murray Darling Basin. Apart from
dropping PDO as a predictor, all other climate indices were included,
allowing the model to select the appropriate predictors.

@@

One of the main difficulties in empirical observation studies related to
the effect of land cover change on rainfall is the lack of continuous
monitoring of land surface variables. Moreover, no specific variable can
possibly be defined that can clearly represent the land surface process.
Given the lack of a full picture of the land surface process, a factor
variable was used in the regression model to represent the abrupt land
surface change (see Equation \eqref{eq:model}). The change could be a
result of either land clearing or bush fires as long as it is permanent
or takes a long time to recover. As indicated we approached the problem
in two different models ways.

In the first method, the tree cover change was used as a predictor in
the regression model, represented by a factor variable \textbf{LC}. The
significance of the coefficient of \textbf{LC}, denoted as \(\beta'_5\)
in Equation \eqref{eq:model}, can be determined by a ratio test.

\begin{equation}
  \mbox{LC} = \left\{
  \begin{array}{ll}
     \mbox{Trees} \\
     \mbox{Removed}
  \end{array} \right.\\
\label{eq:dummyC}
\end{equation}

@@

\begin{equation}
\begin{array}{lll}
g(E(\mathbf{R}_r)) = &\beta'_0 + s'_1(\mathbf{MDB_{monthlyRain}} + \\
  &s'_2(\mathbf{SOI}) + s'_3(\mathbf{IOD}) + \\
  &s'_4(\mathbf{Nino3.4}) + s'_5(\mathbf{Nino4}) + s'_6\mathbf{Season} + \\
  &\beta'_1\mathbf{Trend} + \beta'_2\mathbf{LC} + \boldsymbol{\epsilon'}_r
\end{array}
\label{eq:model2}
\end{equation}

One of the difficulties is to point an exact time to the changes in the
vegetation cover in the two regions. In the QLD region, no exact time
can be assigned to the land clearing. According to the SLATS reports,
the most substantial clearing occurred between 2003 - 2004 . However,
the information on the change in type of land cover during the time
period is missing. Therefore, four scenarios were initially tested in
the analysis, however there was no real difference between these
scenarios. In the NSW/VIC region, severe bush fires were reported in
early January 2003. Hence the ``tree'' cover state was up to December
2002 then it was changed to ``removed'' state from January 2003. As a
starting date, the regression model was run from 1979 for both regions.

@@

To support the regression analysis, a Mann-Whitney Rank-Sum step trend
test was used to detect changes in rainfall as a result of vegetation
cover change. This specific nonparametric statistical test was modified
from the Mann-Whitney U test by Hirsch and Gilroy (1985) and can
identify a step change in data which is cross-correlated. The gridded
rainfall dataset used in this study has a high spatial correlation
between neighbouring pixels due to the underlying interpolation method.
The advantages of using the Rank-Sum test are: (1) it does not depend on
assumptions of the data distribution; (2) it is not restricted to
datasets with no missing data; (3) it is robust and not as easily
influenced by outliers and negative numbers (Hirsch and Gilroy 1985).
However, the test has less power than parametric tests.

The rainfall residuals from the regression model in Equation
\eqref{eq:model2} were used for the step trend test. To detect a step
change, using deseasonalised and detrended data is important (Hirsch and
Gilroy 1985). Furthermore, as rainfall can only be partially attributed
to local sources and conditions, other effects are introduced by large
scale dynamics and changes in climatic factors. The assumption is that
the regression model should remove these effects, and additionally
deseasonalise and detrend the rainfall data. As a result, the local
landuse effects are amplified in the remaining residuals. The test,
described in the following sub section, subsequently associates trends
in the rainfall residuals with tree cover changes.

@@

As indicated, the step trend test is a modified version of the
Mann-Whitney U statistic (Hirsch and Gilroy 1985). As a nonparametric
rank-based test, the Mann-Whitney test does not use the exact values of
rainfall but depends on the ranks of the data. For each month, rainfall
residuals of each year were ranked in an ascending order. The ranking of
January rainfall in a sample pixel k in QLD is illustrated below:

\begin{table}[t]

\caption{\label{tab:exampleRank}Example of ranking rainfall residuals}
\centering
\begin{tabular}{rrr}
\toprule
Year & Rainfall residuals & Rank \$R'\_\{1k\}\$\\
\midrule
1998 & -0.3 & 6\\
1999 & -60.9 & 2\\
2000 & -16.1 & 4\\
2001 & -71.7 & 1\\
2002 & 111.1 & 7\\
\addlinespace
2005 & -7.2 & 5\\
2006 & -60.5 & 3\\
\bottomrule
\end{tabular}
\end{table}

@@

The before and after period in the data formed two groups of samples.
The split point of the two periods was based on the timing of the
vegetation cover changes. In the QLD region, changes occurred anytime
during 2003 and 2004. In contrast to the previous method, the time
period covering the land cover change was excluded, as the nonparametric
test allows missing data. Hirsch and Gilroy (1985) also pointed out that
the power of the test is higher if the data of the change period is
ignored. Hence 2003 and 2004 were excluded from the analysis. As a
result, the post change period was 2005 - 2015 for the Queensland
location.

In the case of NSW/VIC, the bushfires broke out in early January 2003.
The change was within a relatively short period of the year. Therefore
the post change period in this region still started in January 2003.
Following Hirsch and Gilroy (1985), the pre change period was set to
five years (1998 - 2002) in both regions. The length of the post change
period is also difficult for the NSW/VIC region, as the regrowth would
at some point have impacted the local effects.

@@

\begin{equation}
  W_{jk} = \sum_{i=1}^{n_1}R'_{ijk}.
  \label{eq:Wj}
\end{equation}

@@

Instead of completing the Mann-Whitney U-test, Hirsch and Gilroy (1985)
applied a standard normal Z test to the rank-sum statistics. As
highlighted, this modified test accounts for serial temporal
autocorrelation and spatial cross correlation in the data. In the case
here, the deseasonalised and detrended data shows little autocorrelation
in the time series but possesses strong cross correlation between
neighbouring pixels, i.e. \(R>0.99\).

@@

\begin{equation}
  Z'=\frac{\sum_{j=1}^{12}\sum_{k=1}^{ns}W_{jk}-12\cdot ns\cdot\mu_w}{\sqrt{Var(\sum_{j=1}^{12}\sum_{k=1}^{ns}W_{jk})}}.
  \label{eq:Z}
\end{equation}

@@

\begin{table}[t]

\caption{\label{tab:Zscore}The interpretation of Z' score in the step trend test}
\centering
\begin{tabular}{l}
\toprule
 \\
\midrule
\$Z'>0\$  and rainfall decreases after change\\
\$Z'<0\$ and rainfall increases post change\\
\$Z'=0\$ and rainfall does not change\\
\bottomrule
\end{tabular}
\end{table}
